Technique for \earlywarning\ detection of compact binary coalescence and its implications for multi-messenger astronomy

### 2011-06-08 introduction.tex
Made the intro rock your socks.
git-svn-id: https://svn.ligo.caltech.edu/svn/6th-Floor-Papers/lloid_paper@985 ccdd05d5-56c1-461a-bf98-85f6876523b4

As a compact binary system loses energy to gravitational waves (\GW{}s), its
orbital separation decays, leading to a run-away inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges soon
after the innermost stable circular orbit (\ISCO). If a neutron star is
involved it may become tidally disrupted near the merger.  This disrupted
matter can fuel a bright electromagnetic counterpart in the merger's
aftermath~\cite{shibata:2007}.

\GW{} detectors are sensitive the to these \GW{}
signals.  Indeed, the coalescence of compact binary systems consisting of
neutron stars (\textsc{ns}) and/or black holes (\textsc{bh}) is the most
promising source of gravitational radiation for Advanced \LIGO, Virgo, \GEO\,
and \LCGT~\cite{ALIGOWeb, AVirgoWeb, GEOWeb, LCGTWeb}.  Tens of binary
coalescence events per year are expected to be observed in the advanced
detector era later this decade~\cite{Abadie:2010p10836}. In the era of
first-generation detectors the \GW{} community initiated a project
to send alerts when potential \GW{} transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were 30--60 minutes~\cite{HugheyGWPAW2011}, which was an
important achievement, but far too late to catch prompt electromagnetic
emission.

Prompt electromagnetic emission can arise as shells of relativistically
out-flowing matter collide in the inner shock. Such an inner shock from a
compact binary coalescence is believed to be a mechanism for short gamma-ray
bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}. The same inner shocks, or
potentially reverse shocks, can produce a bright accompanying optical
flash~\cite{Sari99}. Prompt emission is a probe into the extreme initial
conditions of the outflow, in contrast with afterglows, which arise in the
external shock with the local medium and are relatively
insensitive to initial conditions. Optical flashes have only been observed to
peak within tens of seconds for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\cite{2008Natur.455..183R}. Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\cite{0004-637X-723-2-1711}. An electromagnetic counterpart would
vastly boost confidence in the \GW{} detection and provide the
tight sky localization necessary to allow determination of the source's host
galaxy, which leads to a redshift measurement. A coincident \GW{}
observation will provide a luminosity distance measurement. With these
together, we can produce precision measurements of the Hubble
constant~\cite{2010ApJ...725..496N}.

To this end, we have the ambition of reporting \GW{} candidates not minutes
\emph{after} the merger, as has already been accomplished, but seconds
\emph{before}.  By examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization
accuracy~\cite{Fairhurst2009} for latency.  Figure \ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming the event rate predictions in~\cite{Abadie:2010p10836}.  The
most likely estimates indicate that 10 sources could be audible (meaning
producing an \SNR\ above 8) 10 seconds or more prior to the merger during one
year of observation.  This number is highly uncertain.  In fact as many as 100
events detected 10 seconds early is plausible. It is also plausible that no
sources will be detected at all.  The gray bands give the pessimistic to
optimistic interval defined in ~\cite{Abadie:2010p10836}. 
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable $t$ seconds before coalescence.  The heavy
solid line is the most likely yearly rate estimate $\dot N_{\mathrm{re}}$
during Advanced \LIGO.  The shaded region represents the confidence interval
$[\dot N_{\mathrm{low}}, \dot N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for detection and
that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a detector in
the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be
detectable within 100 seconds of merger.} 
\end{center}
\end{figure}

>>>>

The coalescence of compact binary systems consisting of neutron stars
 (\textsc{ns}) and/or black holes (\textsc{bh}) is presently the best
understood and most actively pursued target source of gravitational
radiation for ground-based gravitational wave (\GW{}) detectors.
As a compact binary system loses energy to \GW{}s, its
orbital separation decays, leading to a run-away inspiral with the \GW{}
amplitude and frequency increasing until the system eventually merges. At the
merger, our understanding of the system's behavior is incomplete. It is thought
that if a neutron star is involved, it may become tidally disrupted near
the merger and fuel a bright electromagnetic counterpart ~\cite{shibata:2007}.
This counterpart is expected to be rich in information about the source, but 
extracting that information requires rapid identification of candidate
gravitational wave events.

Though compact binary coalescence (\CBC{}) events may be responsible for several
 classes of electromagnetic transients, \CBC{}s are in particular believed to be a
mechanism for short gamma-ray bursts (short \GRB{}s)~\cite{Lee:2005, nakar07}.
In this scenario, prompt electromagnetic emission arises as shells of
 relativistically out-flowing matter collide in the inner shock.
The same inner shocks, or potentially reverse shocks, can produce a bright
 accompanying optical flash~\cite{Sari99}. Prompt emission is a probe into
 the extreme initial conditions of the outflow, in contrast with afterglows,
which arise in the external shock with the local medium and are relatively
insensitive to initial conditions.

Optical flashes typically peak in brightness within tens of seconds and
consequently have only been observed for a handful of long
\GRB{}s~\cite{2011CRPhy..12..255A} by telescopes with extremely rapid response
or, in the case of \textsc{grb 080319b}, by pure serendipity, where several
telescopes were observing a previous \GRB{} in the same
field~\cite{2008Natur.455..183R}. Short \GRB{}s, on the other hand, typically
fade too quickly to even catch the tails of the afterglows. Rapid
\GW{} transient alerts could enable the first observation of
prompt optical flashes and the rise of afterglows from short \GRB{}s, and, for
an exceptional event, perhaps even a glimpse of a
precursor~\cite{0004-637X-723-2-1711}.

Observation of an electromagnetic counterpart to a candidate \GW{} event would
vastly boost confidence in the \GW{} detection and provide the best possible
estimation of the event's positon in the sky localization. With such detailed
information, it will be possible to determine the source's host
galaxy and therefore measure the source's redshift. A coincident \GW{}
observation of this event will also provide a luminosity distance measurement.
Together, these two quantities immediately lead to measurements of the Hubble
constant of unprecedented accuracy ~\cite{2010ApJ...725..496N}.

In the era of first-generation detectors, the \GW{} community initiated a project
to send alerts when potential \GW{} transients were observed in
order to trigger followup observations by optical telescopes.  The best
demonstrated latencies were 30--60 minutes~\cite{HugheyGWPAW2011}, which was an
important achievement, but far too late to catch prompt electromagnetic
emission. In order to extract the maximal science from these events, we have 
the ambition of reporting \GW{} candidates not minutes \emph{after} the merger, 
as has already been accomplished, but seconds \emph{before}.

We show below that by examining the signal to noise ratio (\SNR{}) stream for
threshold crossings before the \GW{} signal leaves the detection band, it is
possible to trade some \SNR{} and sky localization
accuracy~\cite{Fairhurst2009} for latency.  Figure \ref{fig:earlywarning} shows
projected early trigger rates for \textsc{ns}--\textsc{ns} binaries in Advanced
\LIGO\ assuming the event rate predictions in~\cite{Abadie:2010p10836}.  The
most likely estimates indicate that 10 sources could be audible (meaning
producing an \SNR\ above 8) 10 seconds or more prior to the merger during one
year of observation.  This number is highly uncertain.  In fact as many as 100
events detected 10 seconds early is plausible. It is also plausible that no
sources will be detected at all.  The gray bands give the pessimistic to
optimistic interval defined in ~\cite{Abadie:2010p10836}. 
%
\begin{figure}
\begin{center}
\includegraphics{figures/snr_in_time.pdf}
\caption{\label{fig:earlywarning} Expected number of \textsc{ns}--\textsc{ns}
sources that will be detectable $t$ seconds before coalescence.  The heavy
solid line is the most likely yearly rate estimate $\dot N_{\mathrm{re}}$
during Advanced \LIGO.  The shaded region represents the confidence interval
$[\dot N_{\mathrm{low}}, \dot N_{\mathrm{high}}]$ described in
\cite{Abadie:2010p10836}.  Advanced \LIGO\ noise projections are described in
\cite{ALIGONoise}.  Assuming that an \SNR\ of 8 is sufficient for detection and
that we observe $\dot N_{\mathrm{re}} = 40$ events~yr$^{-1}$ with a detector in
the `zero detuning, high power' configuration, about 10~yr$^{-1}$ will be
detectable within 10 seconds before merger and $\sim1$~yr$^{-1}$ will be
detectable within 100 seconds of merger.} 
\end{center}
\end{figure}
